%==============================================================================
%  B.I.O.M.A. — Biologically-Inspired Orchestration of Mutating Agents
%  A dynamic neural-agent framework + a discipline for epistemic honesty
%  in bio-inspired AI, with a pre-registered evaluation protocol.
%
%  Manuscript type: Systems / methods paper in the Registered-Report style.
%  IMPORTANT (integrity): this manuscript reports NO empirical results.
%  It specifies architecture, methodology, and a pre-registered protocol with
%  falsifiable hypotheses. Results are stated as predictions, not findings.
%
%  Compiles with a standard TeX Live / Overleaf install (pdflatex, two passes).
%  Author/affiliation below are placeholders — set them before submission.
%==============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage{caption}

\newcommand{\bioma}{\textsc{B.I.O.M.A.}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Iden}{\mathbf{I}}
\newtheorem{definition}{Definition}
\newtheorem{hypothesis}{Hypothesis}
\newlist{honesty}{description}{1}

\title{\bioma{}: Biologically-Inspired Orchestration of Mutating Agents\\[2pt]
\large A Framework for Dynamic Neural-Agent Orchestration and a Discipline for
Epistemic Honesty in Bio-Inspired AI}

\author{%
Jonathas Cordeiro\thanks{Independent Researcher. Correspondence:
\texttt{jonathas.cordeiro2023@gmail.com}. \emph{Author name and affiliation are
placeholders to be finalized before submission.}}\\
Independent Researcher}

\date{Preprint --- Registered-Report Stage (Stage~1: Design \& Protocol). \\
\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Contemporary ``autonomous agent'' frameworks are predominantly static software
wrappers around a frozen model, and the biological vocabulary frequently
attached to them (``organisms,'' ``evolution,'' ``self-healing'') is rarely
grounded in a concrete mechanism, inviting overclaiming. We present \bioma{}
(Biologically-Inspired Orchestration of Mutating Agents), a PyTorch framework in
which agents are lightweight neural modules organized in a runtime-mutable
directed acyclic graph (DAG) that \emph{expands} and \emph{contracts} during
inference. \bioma{} realizes four capabilities under an explicit
biological-to-engineering mapping: (i)~\emph{mitosis}, function-agnostic runtime
growth by embedding-space clustering plus state-dictionary duplication, in the
Mixture-of-Experts (MoE) / dynamic-network lineage; (ii)~\emph{apoptosis},
structural pruning with knowledge consolidation and deterministic memory
reclamation; (iii)~a \emph{hormonal bus}, an in-process shared-latent blackboard
read by cosine attention with eventual-consistency semantics; and (iv)~an
\emph{energy homeostasis} model, a non-conservative compute budget with
uncertainty regularization. Our second and equally central contribution is
methodological: a three-level \emph{metaphor$\rightarrow$mechanism} honesty
taxonomy (N1 concrete / N2 analogy / N3 narrative) that disciplines every claim
and directly targets the field's chronic overclaiming problem. We formalize
falsifiable, operational metrics---most notably a statistically defined
Computational Survival Rate (CSR) with Wilson lower bounds---and pre-register a
$2\times2$ factorial ablation on a synthetic benchmark with analytic ground
truth, together with an identifiability analysis that ties expected benefit to a
controllable coupling parameter. \textbf{This manuscript is a Stage-1
registered report: it contains no empirical results.} We state hypotheses,
analysis, and stopping rules in advance, and we commit to publishing refutations.
We conclude with an optional evolutionary auto-refactoring extension whose
security and reproducibility boundaries we make explicit.
\end{abstract}

\medskip
\noindent\textbf{Keywords:} dynamic neural networks, mixture-of-experts,
multi-agent systems, artificial life, neuroevolution, blackboard architectures,
epistemic honesty, registered reports, reproducibility.

%==============================================================================
\section{Introduction}
%==============================================================================
The dominant pattern for ``agentic'' AI couples a fixed neural network to an
orchestration layer that routes text through tools and other model instances.
The topology of computation is static: the same parameters and the same graph
serve a trivial prompt and a multi-domain crisis alike. In parallel, a large and
productive body of work has shown that \emph{the graph itself} can be a
first-class, adaptive object---networks can grow while preserving function
\cite{chen2016net2net,rusu2016progressive,yoon2018den}, route computation
conditionally through specialized experts \cite{shazeer2017moe,fedus2022switch},
be searched or evolved as architectures
\cite{zoph2017nas,elsken2019nas,real2019amoeba,stanley2002neat}, and coordinate
through shared external memory rather than message passing
\cite{graves2016dnc,hayesroth1985blackboard}. \bioma{} asks a simple question:
\emph{what does an agent framework look like if the population of agents and the
communication substrate are themselves dynamical objects, created and destroyed
in response to the cognitive load of the input?}

Bio-inspiration is a double-edged sword. It is a rich source of mechanisms rooted in the artificial-life tradition
\cite{langton1989alife}---%
morphogenesis \cite{mordvintsev2020growing}, stigmergy \cite{dorigo1996ant},
programmed cell death, neuroevolution
\cite{stanley2019neuroevolution}---but it is also a well-documented vector for
scholarship that conflates a suggestive name with a demonstrated property
\cite{lipton2019troubling}. A system that ``undergoes mitosis'' and
``apoptosis'' is easy to \emph{narrate} and hard to \emph{falsify} unless every
term is pinned to a measurable operation. We therefore treat epistemic honesty
not as a disclaimer but as a design constraint with its own artifacts and
acceptance tests.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt]
\item \textbf{A framework} (\S\ref{sec:framework}) for dynamic neural-agent
orchestration: runtime DAG mitosis, apoptosis with deterministic memory
reclamation, an in-process hormonal bus, and a non-conservative energy
homeostasis model---each specified with contracts and gradient-safety
invariants.
\item \textbf{A methodological discipline} (\S\ref{sec:honesty}): a three-level
metaphor$\rightarrow$mechanism taxonomy that is applied uniformly and is itself
subject to acceptance criteria (e.g., zero unqualified claims of ``life'').
\item \textbf{A pre-registered evaluation protocol} (\S\ref{sec:protocol}):
operational, falsifiable metrics; a $2\times2$ factorial ablation on a synthetic
benchmark with analytic ground truth; an identifiability analysis; a statistical
analysis plan with power, corrections, and stopping rules.
\item \textbf{An optional evolutionary extension} (\S\ref{sec:evo}) for
automated code refactoring, with explicit security (\texttt{subprocess}
isolation, never \texttt{exec}) and reproducibility boundaries.
\end{enumerate}

\paragraph{Scope and honesty statement.}
This is a Stage-1 registered report. We deliberately \emph{do not} report
performance numbers, survival rates, or convergence curves, because none have
been measured; the reference implementation, built on PyTorch \cite{paszke2019pytorch}, targets
a CPU-only commodity host
and the full protocol has not been executed. Presenting hypothetical numbers as
findings would be data fabrication and is antithetical to the paper's thesis.
What we commit to here---the architecture, the honesty discipline, the metrics,
and the analysis plan---is fixed in advance so that a future Stage-2 report can
be judged against predictions rather than post-hoc rationalizations
\cite{chambers2013registered,pineau2021repro}.

%==============================================================================
\section{Related Work}
%==============================================================================
\paragraph{Dynamic and growing networks.}
Net2Net \cite{chen2016net2net} introduced function-preserving transforms that
widen or deepen a network while exactly preserving its function at the moment of
expansion; progressive networks \cite{rusu2016progressive} and dynamically
expandable networks \cite{yoon2018den} grow capacity for new tasks. HyperNetworks \cite{ha2017hypernetworks} take a
complementary route, generating a network's weights from a smaller meta-network.
\bioma{}
inherits the \emph{idea} of runtime structural change but---crucially---its
duplication is not a Net2Net identity guarantee; it is a state-dictionary
deep-copy whose post-split behavioral equality is a \emph{copy sanity check},
not an algebraic width-expansion identity (see \S\ref{sec:framework}).

\paragraph{Conditional computation and Mixture-of-Experts.}
Sparsely-gated MoE \cite{shazeer2017moe} and Switch Transformers
\cite{fedus2022switch} activate input-dependent sub-networks through a learned
router with load-balancing. \bioma{}'s ``stem cell'' is a router that decides
\emph{whether and how much to expand}; its mitosis is MoE-style specialization
by clustering, and we credit that lineage explicitly rather than claiming a
novel growth primitive.

\paragraph{Architecture search and neuroevolution.}
NAS \cite{zoph2017nas,elsken2019nas} frames topology as a search problem;
regularized evolution \cite{real2019amoeba} and NEAT
\cite{stanley2002neat,stanley2019neuroevolution} evolve architectures and
weights. Our optional evolutionary extension (\S\ref{sec:evo}) sits in the
genetic-improvement tradition \cite{koza1992gp,petke2018gi,legoues2012genprog}
applied to source code, not to network weights.

\paragraph{Morphogenesis versus routed growth.}
Growing Neural Cellular Automata \cite{mordvintsev2020growing} are the most
honest artificial-life reference for \emph{growth}, but they are a
\emph{distinct} paradigm: emergent self-organization from a homogeneous local
rule. \bioma{} grows by \emph{central router decision}, not emergence; we cite
NCA as a contrast and reserve the words ``morphogenesis'' and ``emergence'' for
explicitly speculative future work.

\paragraph{Communication substrates.}
Rather than text-API message passing, \bioma{} uses a shared latent field read
by attention---an in-process descendant of blackboard architectures
\cite{hayesroth1985blackboard}, stigmergy \cite{dorigo1996ant}, and
differentiable external memory \cite{graves2016dnc,vaswani2017attention}. We do
\emph{not} claim ``real-time, overhead-free'' communication; we claim the
elimination of serialization cost under eventual-consistency semantics.

\paragraph{Pruning, distillation, and adaptive compute.}
Apoptosis draws on classical pruning \cite{lecun1990obd,frankle2019lottery} and
knowledge distillation \cite{hinton2015distill}; the energy budget is in the
spirit of adaptive computation time / ponder cost \cite{graves2016act}. None of
these are our invention; our contribution is their \emph{orchestrated
integration} under one honest lifecycle.

\paragraph{Reproducibility and scholarship norms.}
We adopt community reproducibility guidance
\cite{pineau2021repro,gundersen2018repro}, the registered-report format
\cite{chambers2013registered}, and take seriously the critique of overclaiming
in ML scholarship \cite{lipton2019troubling}. The CPU-only target is a
deliberate nod to the hardware-lottery argument \cite{hooker2021hardware}: a
method whose value depends on scarce accelerators is harder to reproduce and to
trust.

%==============================================================================
\section{The \bioma{} Framework}
\label{sec:framework}
%==============================================================================
\bioma{} is organized as a layered system around five modules:
\texttt{organism\_core} (the passive neural unit), \texttt{mitosis\_engine} (the
fission controller), \texttt{hormonal\_bus} (the shared latent substrate),
\texttt{simulation\_harness} (the benchmark and judge), and an optional
\texttt{evolutionary\_coder} (\S\ref{sec:evo}). We summarize the design
principles and then each mechanism, giving the mathematical contract without
implementation detail.

\subsection{Design principles}
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt]
\item \textbf{Dual naming.} Every entity carries a biological name \emph{and} an
engineering name; the engineering name is authoritative for code, logs, and
metrics.
\item \textbf{Gradient safety by isolation.} Structural mutation occurs only
between backward passes, never concurrently with one; children are born
\emph{detached}, with disjoint autograd graphs.
\item \textbf{Determinism where it counts.} Validation runs are executed
serially for \emph{logical} reproducibility; bit-exact numerical reproducibility
is documented as best-effort only.
\item \textbf{Device honesty.} The reference target is CPU-only;
\texttt{empty\_cache} is guarded by \texttt{cuda.is\_available()}; no result
depends on a specific accelerator.
\end{enumerate}

\subsection{Organism core: genome, mutation, energy}
An organism is an \texttt{nn.Module} $\theta$ augmented with a \emph{genome}: a
tuple $(\mathrm{id}, g, \mathrm{parent}, s, \sigma, b, \mathrm{dev}, h)$ of unique
id, generation $g$, parent id, specialization label $s$, mutation seed $\sigma$,
birth step $b$, device, and an architecture signature $h$ (a canonical hash of
topology, shapes, and dtypes). Replication is a deep-copy of the state
dictionary; independence is asserted by distinct storage pointers. Mutation
perturbs each layer's weights with layer-relative Gaussian noise,
\begin{equation}
\theta'_\ell \;=\; \theta_\ell + \varepsilon_\ell,\qquad
\varepsilon_\ell \sim \mathcal{N}\!\big(0,\; (m\cdot(\mathrm{std}(\theta_\ell)+\epsilon))^2\big),
\end{equation}
drawn from a dedicated \texttt{torch.Generator} for reproducibility, with a
mutability mask freezing normalization parameters and a finiteness rejection
step. Energy accounting (\S\ref{sec:energy}) is attached as a forward hook under
\texttt{no\_grad}, so it never perturbs gradients.

\subsection{Mitosis: semantic divergence and clustered fission}
Given an incoming scenario embedded by a shared encoder $\Phi$ into
$\{x_i\}\subset\R^d$, the stem-cell router computes a scalar \emph{semantic
divergence}
\begin{equation}
D \;=\; \sum_{k=1}^{4} w_k\, s_k, \qquad
\textstyle\sum_k w_k = 1,\quad s_k\in[0,1],
\label{eq:divergence}
\end{equation}
a convex combination of four normalized signals: (i)~the entropy of the mass
distribution across candidate clusters (\emph{cluster-mass entropy}), (ii)~a
silhouette score \cite{rousseeuw1987silhouettes} computed \emph{on the candidate
partition} (not before it, to avoid circularity), (iii)~an inter/intra cluster
variance ratio, and (iv)~mean pairwise distance. Fission fires under
\emph{hysteresis}: the population expands when $D$ exceeds an upper threshold
$\tau_{\uparrow}$ (with an exponential moving average and a cooldown), and the
decision to contract uses a lower threshold $\tau_{\downarrow}<\tau_{\uparrow}$,
preventing oscillation. The number of children $k^\star$ is chosen by scanning
$k\in\{2,\dots,k_{\max}\}$ and maximizing $\mathrm{Quality}(k)-\lambda k$;
if no $k>1$ beats $k=1$, \emph{no mitosis occurs}. Each child inherits the
parent's weights (deep-copy) plus a specialization bias toward its cluster
centroid, injected through a fixed-dimensional conditioning context rather than
by destructively overwriting weights. Children are inserted as DAG leaves with
disjoint autograd graphs.

\begin{quotation}
\noindent\textbf{Honesty note.} The post-split equality
$\lVert f_{\text{after}}-f_{\text{before}}\rVert<10^{-5}$ (before any adaptation)
is a \emph{deep-copy sanity check}, \emph{not} the Net2Net function-preservation
identity \cite{chen2016net2net}. We additionally require an \emph{effective
differentiation} test after adaptation, so that mitosis is shown to do work
rather than to be expensive theater.
\end{quotation}

\subsection{Hormonal bus: shared latent field with attention reads}
The bus is a tensor $\mathbf{M}\in\R^{S\times d}$ of $S$ slots, one per live
cell. A cell \emph{secretes} by an exponential-moving-average write with norm
clamping and non-finite sanitization; it \emph{senses} by true cosine attention
over occupied, non-stale slots:
\begin{equation}
a_j \;=\; \mathrm{softmax}_j\!\left(\frac{1}{\tau}\,
\frac{\langle q, k_j\rangle}{\lVert q\rVert\,\lVert k_j\rVert}\right),
\qquad
r \;=\; \sum_{j} a_j\, v_j,
\label{eq:attn}
\end{equation}
where $q$ is the reader's query and $(k_j,v_j)$ are slot key/value. Temporal
dissipation is applied at each \texttt{tick} to all occupied slots so that stale
signals decay and attention can discard them. Concurrency uses versioned
snapshot isolation (copy under lock, compute outside it) and an explicit
differentiability policy (non-differentiable by default; a stop-gradient on keys
in the opt-in differentiable mode).

\begin{quotation}
\noindent\textbf{Honesty note.} Communication is \emph{eventually consistent}: a
read observes the last consolidated tick, not the instant a peer wrote. The gain
over text-API agents is the elimination of serialization, at the cost of lock
synchronization and tick latency.
\end{quotation}

\subsection{Energy homeostasis}
\label{sec:energy}
Each cell carries a scalar state $E_i$ in a \texttt{float64} buffer updated under
\texttt{no\_grad}:
\begin{equation}
E_i(t{+}1)=\mathrm{clamp}\!\Big(
E_i(t) - \lambda_{\text{basal}} E_i(t) - \Delta E^{\text{work}}
- \underbrace{\kappa\,H_{\text{norm}}}_{\Delta E^{\text{unc}}}
+ \eta\,\max(0,\Delta \mathrm{util}),\;
E_{\min}, E_{\max}\Big).
\label{eq:energy}
\end{equation}
Here $\Delta E^{\text{work}}$ is an analytic FLOP/token proxy;
$H_{\text{norm}}=H/\log K$ is the Shannon uncertainty of the output normalized
by $\log K$; and regeneration $\eta\max(0,\Delta\mathrm{util})$ rewards
\emph{progress} (sub-domain loss reduction), never mere activity.

\begin{quotation}
\noindent\textbf{Honesty note.} The sources, sinks, and \texttt{clamp} in
Eq.~\eqref{eq:energy} make the system \emph{non-conservative by design}. The
auditable property is therefore \emph{accounting consistency} (no double
counting, no floating-point drift, $E_i\ge0$), not physical conservation.
``Entropy'' here is Shannon uncertainty (a chosen regularizer), not
thermodynamic entropy.
\end{quotation}

\subsection{Apoptosis and lifecycle}
Termination is a state machine
$\textsc{alive}\!\to\!\textsc{dying}\!\to\!\textsc{transferring}\!\to\!\textsc{dead}$
triggered by an OR of four conditions with hysteresis (except energy
exhaustion, which is immediate): (1)~task solved (loss $<\tau_{\text{loss}}$),
(2)~energy exhausted ($E_i\le E_{\text{death}}$), (3)~low marginal contribution,
(4)~senescence (age $>T_{\max}$). Before release, the cell transfers its
representation to the parent by utility-weighted interpolation (LERP/SLERP) or
embedding distillation \cite{hinton2015distill}, with a pre-merge snapshot for
rollback if global loss rises. Cleanup follows a canonical order: drop
references $\to$ detach $\to$ move to CPU $\to$ \texttt{gc.collect()} $\to$
\texttt{empty\_cache()} (CUDA-guarded), verified by a leak auditor. A topological
scheduler ensures a parent reduces only after all children have apoptosed.

\subsection{Concurrency and gradient safety}
The reference concurrency model is \emph{cooperative single-thread}: bus
coordination runs on the event loop, and the only true parallelism is per-cell
computation in an executor over disjoint tensors. A budget semaphore admits new
cells (resolving a cooperative reservation race), a no-\texttt{await}-under-lock
rule and a total lock order prevent deadlock, and an ownership assertion before
each backward enforces autograd-graph disjointness. We state plainly that under
the Python GIL, single-process parallelism is bounded by Amdahl's law; we prefer
batching sibling cells over adding workers.

%==============================================================================
\section{A Metaphor$\rightarrow$Mechanism Discipline}
\label{sec:honesty}
%==============================================================================
Our methodological contribution is a uniform discipline for naming. Every term
is assigned an engineering mechanism, a technical family, and a
\emph{honesty level}:
\begin{honesty}
\item[\textbf{N1 (concrete)}] a verifiable mechanism with a measurable operation.
\item[\textbf{N2 (analogy)}] a defensible functional analogy, not a literal claim.
\item[\textbf{N3 (narrative)}] creative license with no mechanical referent;
permitted only when explicitly flagged.
\end{honesty}

Table~\ref{tab:dict} is the canonical dictionary and the single source of truth
for terminology. It is not decoration: the protocol (\S\ref{sec:protocol})
includes an acceptance test requiring \emph{zero} unqualified claims of ``life,''
``consciousness,'' ``emergence,'' or ``thermodynamics.''

\begin{table}[t]
\centering
\caption{Canonical metaphor$\rightarrow$mechanism dictionary. The engineering
name is authoritative.}
\label{tab:dict}
\small
\begin{tabular}{@{}p{2.7cm}p{6.4cm}p{3.0cm}c@{}}
\toprule
\textbf{Biological term} & \textbf{Engineering mechanism} & \textbf{Family} & \textbf{Level}\\
\midrule
Mitosis & Clustered spawning + state-dict deep-copy + conditional routing & MoE / growing nets & N1\\
Stem cell & Router measuring embedding dispersion; decides expansion & MoE gating / NAS & N1\\
Semantic divergence & Scalar statistic over embedding geometry, Eq.~\eqref{eq:divergence} & Information theory / clustering & N1\\
Homeostasis & Regulation of an internal variable to a viable band & Control / resource accounting & N2\\
Energy & Scalar budget counter (FLOP/token proxy) + uncertainty penalty & Conditional compute & N2\\
Entropy (drain) & Shannon uncertainty regularizer of the output & Entropy regularization & N2\\
Apoptosis & Structural pruning + consolidation + memory reclamation & Pruning + distillation + GC & N1\\
Hormone / exudation & EMA write + cosine-attention read on shared tensor, Eq.~\eqref{eq:attn} & Blackboard / stigmergy / attention & N1\\
Reverse distillation & Promotion of a winning transform to a reusable library & Pattern mining / library update & N2\\
``Life'' / ``consciousness'' & --- (no mechanical referent) & --- & \textbf{N3: forbidden unqualified}\\
\bottomrule
\end{tabular}
\end{table}

We regard this table as the paper's most transferable artifact: it is a template
any bio-inspired system can adopt to convert suggestive names into auditable
claims, and it turns ``did the authors overclaim?'' from a matter of taste into a
checkable property.

%==============================================================================
\section{Pre-Registered Evaluation Protocol}
\label{sec:protocol}
%==============================================================================
We now fix, in advance, the benchmark, the metrics, the experimental design, the
statistical plan, and the hypotheses. No aspect of this section may be altered
post hoc without disclosure.

\subsection{Synthetic benchmark with analytic ground truth}
We use a controllable multi-domain generator. For $K$ latent domains we draw
orthonormal prototypes $\mathbf{P}\in\R^{K\times d}$ via a QR factorization of a
Gaussian matrix, generate point clouds $x=p_k+\sigma\varepsilon$, and define a
domain-coupling matrix $\mathbf{C}$ normalized so that the spectral radius
$\rho(\gamma\mathbf{C})<1$. The \emph{analytic} global solution is
\begin{equation}
\mathbf{S}^\star \;=\; (\Iden - \gamma\mathbf{C})^{-1}\mathbf{P},
\label{eq:groundtruth}
\end{equation}
held \emph{detached} and never observed by any organism. The coupling parameter
$\gamma\in[0,\rho^{-1})$ controls how much the domains must be solved
\emph{jointly}: at $\gamma\to0$ the domains are independent and a monolithic
solver suffices; as $\gamma$ grows, coordination becomes necessary.

\subsection{Operational metrics}
\paragraph{Computational Survival Rate (CSR).}
For a cell $c$ define the predicate
\begin{equation}
\textsc{survive}(c) = \mathrm{healthy\_fwd} \wedge \neg\mathrm{nan\_inf}
\wedge \mathrm{transfer\_ok} \wedge \mathrm{clean\_apoptosis}
\wedge \neg\mathrm{leak} \wedge \neg\mathrm{race},
\end{equation}
and $\mathrm{CSR}=\frac{1}{N}\sum_{c}\mathbf{1}[\textsc{survive}(c)]$ over $N$
declared births. Because CSR is a proportion over a stochastic process, the
acceptance criterion is \emph{statistical}: the Wilson \cite{wilson1927interval}
lower bound of CSR at confidence $1-\alpha$ must exceed a pre-declared threshold,
\emph{or} CSR $=1$ conditioned on a second, tolerance-free oracle (the
\texttt{gc}-object cross-check) classifying every sub-unity termination as
genuine necrosis. ``Survival'' is explicitly \emph{not} the absence of
apoptosis: apoptosis is healthy and expected.

\paragraph{Convergence, decision fidelity, identifiability.}
Convergence is $\cos(\hat{\mathbf{s}}, \mathbf{S}^\star)$ for the root synthesis
$\hat{\mathbf{s}}$. Mitosis decision fidelity is reported as a \emph{confusion
matrix}: firing on multi-domain scenarios and \emph{suppressing} on uni-domain
scenarios, with precision/recall above a declared floor. For identifiability we
report the trivial baseline $\cos(\hat{\mathbf{s}}, \bar{\mathbf{p}})$ (against
the prototype mean) alongside $\cos(\hat{\mathbf{s}}, \mathbf{S}^\star)$ and
sweep $\gamma$; if both are high the scenario is trivial and is reported as such.

\paragraph{Memory-leak criterion (CPU-only aware).}
Rather than ``return to baseline $\pm1\%$'' (fragile under allocator noise), the
leak criterion is the \emph{absence of an upward trend} in a soak test
(regression slope $b\le b_{\max}$ with $p\ge0.05$) plus a tolerance-free
\texttt{gc} count of live \texttt{NeuralOrganism} objects returning to zero.

\subsection{Experimental design}
The core design is a $2\times2$ factorial: mitosis $\in\{\text{on},\text{off}\}$
$\times$ hormonal bus $\in\{\text{on},\text{off}\}$, with $R$ replicates over a
seed grid and fixed data and hyperparameters. Control conditions include a
monolithic solver, a communication-free swarm, and---critically---%
\emph{adversarial scenarios in which splitting is wrong} (coupled domains that
require a joint representation), so that the benchmark cannot be won by
indiscriminate fission.

\subsection{Statistical analysis plan}
We pre-commit to: Welch's $t$-test or Mann--Whitney per normality checks; effect
sizes (Cohen's $d$ / Cliff's $\delta$) with $95\%$ bootstrap confidence
intervals; family-wise error control via Holm--Bonferroni
\cite{holm1979bonferroni}; and an a-priori power analysis to size $R$ for
power $\ge0.8$ at a pre-declared effect size. Validation runs use
\texttt{max\_workers}$=1$ (serial, logically deterministic); performance and
correctness are measured separately. Metric tolerances are derived empirically
from CPU floating-point jitter (typically $10^{-4}$--$10^{-3}$, not $10^{-5}$).

\subsection{Hypotheses}
We state falsifiable predictions and pre-commit to publishing refutations.
\begin{hypothesis}[Mitosis utility]
On multi-domain scenarios, enabling mitosis increases
$\cos(\hat{\mathbf{s}},\mathbf{S}^\star)$ over the monolithic baseline, and the
margin \emph{increases} with coupling $\gamma$.
\end{hypothesis}
\begin{hypothesis}[Bus utility]
Enabling the hormonal bus increases convergence over a communication-free swarm,
with the largest effect at high $\gamma$.
\end{hypothesis}
\begin{hypothesis}[Decision fidelity]
The mitosis router fires on multi-domain and suppresses on uni-domain scenarios
with precision and recall above the declared floor.
\end{hypothesis}
\begin{hypothesis}[Survival]
The Wilson lower bound of CSR meets the declared threshold with no unrecovered
anomaly and no memory-leak trend.
\end{hypothesis}
\begin{hypothesis}[Effective differentiation]
After adaptation, sibling cells diverge and specialize (inter-sibling distance
and per-domain loss reduction exceed a floor), i.e., mitosis performs work.
\end{hypothesis}
A null result on H1 or H2---mitosis or communication failing to help---is a
valid, publishable outcome and would itself constrain the design space.

%==============================================================================
\section{Optional Extension: Evolutionary Auto-Refactoring}
\label{sec:evo}
%==============================================================================
As a capability extension built atop the frozen framework, the
\texttt{evolutionary\_coder} applies genetic-improvement search
\cite{koza1992gp,petke2018gi,legoues2012genprog} to refactor a target Python
function, \emph{reusing} the orchestration, lifecycle, and telemetry machinery
but operating on \emph{programs, not tensors}. Four honesty corrections are
mandatory and are themselves acceptance criteria:
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt]
\item \textbf{No gradients.} Refactoring is discrete, non-differentiable search;
``dropped gradients during dynamic execution'' is a category error. Static audit
forbids any autograd reference in this module.
\item \textbf{Evolution does not discover a complexity class.} Turning an
$O(N^2)$ sort into $O(N\log N)$ does not emerge from random mutation; the gain
comes either from an AST transform \emph{already present in the catalog} or from
an LLM proposal (knowledge, not natural selection). Each variant is
\emph{labeled by origin}, and success is measured as latency reduction against a
baseline, never narrated as emergent discovery.
\item \textbf{\texttt{exec} does not isolate.} Variants run \emph{only} in a
resource-limited \texttt{subprocess} (\texttt{setrlimit} / Windows Job Objects,
hard timeout, no network); in-process \texttt{exec}/\texttt{eval} of untrusted
code is forbidden.
\item \textbf{``Reverse distillation'' is pattern harvesting.} The winning
transform is promoted to the deterministic catalog; no network ``learns the
concept of clean code.''
\end{enumerate}
Fitness is \emph{lexicographic}: correctness as a hard gate (any failing test
$\Rightarrow$ fitness $0$), then median latency over $R$ repetitions, then memory
delta. Two separated paths preserve the reproducibility decision of the
framework: an AST-deterministic path (in the validated set) and a
quarantined, explicitly non-reproducible LLM path (excluded from statistical
validation).

%==============================================================================
\section{Threats to Validity}
%==============================================================================
\paragraph{Construct validity.}
``Energy'' is a compute proxy, not a physical quantity; biological terms are
scaffolds (Table~\ref{tab:dict}). Semantic divergence depends on the encoder
$\Phi$; a poor $\Phi$ can mask or fabricate multi-domain structure. We freeze
$\Phi$'s contract in advance and report sensitivity.

\paragraph{Internal validity.}
Asynchronous execution is a source of nondeterminism; we neutralize it for
validation by serial execution and a canonical reduction order, and we
audit for read-after-write races directly.

\paragraph{External validity.}
The benchmark is synthetic, the models are small, and the host is CPU-only.
Results---if obtained---may not transfer to large-scale or GPU regimes or to
real workloads. We make no scaling claims beyond the tested regime, in keeping
with the hardware-lottery caution \cite{hooker2021hardware}.

\paragraph{Statistical validity.}
Multiple comparisons are corrected \cite{holm1979bonferroni}; replicate counts
are sized by power analysis; tolerances are empirical, not aspirational.

%==============================================================================
\section{Ethical and Epistemic Considerations}
%==============================================================================
The primary ethical risk of this line of work is not physical but
\emph{rhetorical}: presenting a Mixture-of-Experts router with pruning as an
``autonomous digital life form'' misleads readers and erodes trust in the field
\cite{lipton2019troubling}. Our N1/N2/N3 discipline and its zero-unqualified-%
claim acceptance test are a direct, checkable countermeasure. The optional
evolutionary extension executes generated code; we contain this strictly in an
isolated \texttt{subprocess} with resource limits and no network, and we treat
all generated code as untrusted. We foresee no dual-use concern beyond standard
code-execution hygiene.

%==============================================================================
\section{Conclusion and Future Work}
%==============================================================================
\bioma{} reframes an agent framework as a dynamical system: a population of
neural modules that grows and prunes itself in a runtime-mutable DAG,
coordinated through a shared latent field and bounded by a compute budget. Its
mechanisms are not new in isolation---they are borrowed, with credit, from MoE,
dynamic networks, blackboard systems, pruning, and distillation---but their
honest integration under a single lifecycle, and the discipline that keeps the
biological narrative from outrunning the engineering, are the contribution.
Equally, the pre-registered protocol commits us to be judged by predictions. The
natural next step is the Stage-2 report: execute the $2\times2$ ablation and the
soak test on the reference implementation, and report---whatever the outcome---%
the CSR, the convergence-versus-$\gamma$ curves, and the mitosis confusion
matrix, including any refutation of H1--H5. Longer-term directions include a
genuinely emergent (NCA-style) growth rule as an explicitly labeled research
branch, and a GPU study to test whether the orchestration survives contact with
real parallelism.

%==============================================================================
\section*{Reproducibility Statement}
%==============================================================================
All hypotheses, metrics, and analyses are fixed in this Stage-1 document.
Planned artifacts for Stage~2 include: seeded, versioned configuration; a
declared births denominator $N$; serial validation for logical reproducibility;
the synthetic-benchmark generator with its coupling sweep; and the leak-auditor
and race-probe harnesses. Bit-exact numerical reproducibility is documented as
best-effort under serial execution only. We will release code and logs with the
Stage-2 report.

%==============================================================================
\section*{Data Availability}
%==============================================================================
No datasets were used or generated for this Stage-1 report; the Stage-2
benchmark is fully synthetic and generated from released seeds.

%==============================================================================
\begin{thebibliography}{99}
\small
\bibitem{paszke2019pytorch} A.~Paszke et al.,
``PyTorch: An Imperative Style, High-Performance Deep Learning Library,''
\emph{NeurIPS}, 2019.
\bibitem{shazeer2017moe} N.~Shazeer et al.,
``Outrageously Large Neural Networks: The Sparsely-Gated Mixture-of-Experts
Layer,'' \emph{ICLR}, 2017.
\bibitem{fedus2022switch} W.~Fedus, B.~Zoph, N.~Shazeer,
``Switch Transformers: Scaling to Trillion Parameter Models with Simple and
Efficient Sparsity,'' \emph{JMLR}, 2022.
\bibitem{chen2016net2net} T.~Chen, I.~Goodfellow, J.~Shlens,
``Net2Net: Accelerating Learning via Knowledge Transfer,'' \emph{ICLR}, 2016.
\bibitem{rusu2016progressive} A.~A.~Rusu et al.,
``Progressive Neural Networks,'' arXiv:1606.04671, 2016.
\bibitem{yoon2018den} J.~Yoon, E.~Yang, J.~Lee, S.~J.~Hwang,
``Lifelong Learning with Dynamically Expandable Networks,'' \emph{ICLR}, 2018.
\bibitem{zoph2017nas} B.~Zoph, Q.~V.~Le,
``Neural Architecture Search with Reinforcement Learning,'' \emph{ICLR}, 2017.
\bibitem{elsken2019nas} T.~Elsken, J.~H.~Metzen, F.~Hutter,
``Neural Architecture Search: A Survey,'' \emph{JMLR}, 2019.
\bibitem{ha2017hypernetworks} D.~Ha, A.~Dai, Q.~V.~Le,
``HyperNetworks,'' \emph{ICLR}, 2017.
\bibitem{mordvintsev2020growing} A.~Mordvintsev, E.~Randazzo, E.~Niklasson,
M.~Levin, ``Growing Neural Cellular Automata,'' \emph{Distill}, 2020.
\bibitem{hayesroth1985blackboard} B.~Hayes-Roth,
``A Blackboard Architecture for Control,'' \emph{Artificial Intelligence},
26(3), 1985.
\bibitem{dorigo1996ant} M.~Dorigo, V.~Maniezzo, A.~Colorni,
``Ant System: Optimization by a Colony of Cooperating Agents,''
\emph{IEEE Trans. Systems, Man, and Cybernetics--B}, 26(1), 1996.
\bibitem{graves2016dnc} A.~Graves et al.,
``Hybrid Computing Using a Neural Network with Dynamic External Memory,''
\emph{Nature}, 538, 2016.
\bibitem{vaswani2017attention} A.~Vaswani et al.,
``Attention Is All You Need,'' \emph{NeurIPS}, 2017.
\bibitem{graves2016act} A.~Graves,
``Adaptive Computation Time for Recurrent Neural Networks,''
arXiv:1603.08983, 2016.
\bibitem{stanley2002neat} K.~O.~Stanley, R.~Miikkulainen,
``Evolving Neural Networks through Augmenting Topologies,''
\emph{Evolutionary Computation}, 10(2), 2002.
\bibitem{real2019amoeba} E.~Real, A.~Aggarwal, Y.~Huang, Q.~V.~Le,
``Regularized Evolution for Image Classifier Architecture Search,''
\emph{AAAI}, 2019.
\bibitem{stanley2019neuroevolution} K.~O.~Stanley, J.~Clune, J.~Lehman,
R.~Miikkulainen, ``Designing Neural Networks through Neuroevolution,''
\emph{Nature Machine Intelligence}, 1, 2019.
\bibitem{koza1992gp} J.~R.~Koza,
\emph{Genetic Programming: On the Programming of Computers by Means of Natural
Selection}, MIT Press, 1992.
\bibitem{legoues2012genprog} C.~Le~Goues, T.~Nguyen, S.~Forrest, W.~Weimer,
``GenProg: A Generic Method for Automatic Software Repair,''
\emph{IEEE Trans. Software Engineering}, 38(1), 2012.
\bibitem{petke2018gi} J.~Petke et al.,
``Genetic Improvement of Software: A Comprehensive Survey,''
\emph{IEEE Trans. Evolutionary Computation}, 22(3), 2018.
\bibitem{lipton2019troubling} Z.~C.~Lipton, J.~Steinhardt,
``Troubling Trends in Machine Learning Scholarship,''
\emph{ACM Queue} / arXiv:1807.03341, 2019.
\bibitem{chambers2013registered} C.~D.~Chambers,
``Registered Reports: A New Publishing Initiative at Cortex,''
\emph{Cortex}, 49(3), 2013.
\bibitem{pineau2021repro} J.~Pineau et al.,
``Improving Reproducibility in Machine Learning Research,'' \emph{JMLR}, 2021.
\bibitem{gundersen2018repro} O.~E.~Gundersen, S.~Kjensmo,
``State of the Art: Reproducibility in Artificial Intelligence,''
\emph{AAAI}, 2018.
\bibitem{langton1989alife} C.~G.~Langton (ed.),
\emph{Artificial Life}, Addison-Wesley, 1989.
\bibitem{hooker2021hardware} S.~Hooker,
``The Hardware Lottery,'' \emph{Communications of the ACM}, 64(12), 2021.
\bibitem{wilson1927interval} E.~B.~Wilson,
``Probable Inference, the Law of Succession, and Statistical Inference,''
\emph{J. American Statistical Association}, 22, 1927.
\bibitem{holm1979bonferroni} S.~Holm,
``A Simple Sequentially Rejective Multiple Test Procedure,''
\emph{Scandinavian J. Statistics}, 6(2), 1979.
\bibitem{rousseeuw1987silhouettes} P.~J.~Rousseeuw,
``Silhouettes: A Graphical Aid to the Interpretation and Validation of Cluster
Analysis,'' \emph{J. Computational and Applied Mathematics}, 20, 1987.
\bibitem{lecun1990obd} Y.~LeCun, J.~S.~Denker, S.~A.~Solla,
``Optimal Brain Damage,'' \emph{NeurIPS}, 1990.
\bibitem{frankle2019lottery} J.~Frankle, M.~Carbin,
``The Lottery Ticket Hypothesis: Finding Sparse, Trainable Neural Networks,''
\emph{ICLR}, 2019.
\bibitem{hinton2015distill} G.~Hinton, O.~Vinyals, J.~Dean,
``Distilling the Knowledge in a Neural Network,'' arXiv:1503.02531, 2015.
\end{thebibliography}

%==============================================================================
\appendix
\section{Notation}
%==============================================================================
\begin{table}[h]
\centering
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Symbol} & \textbf{Meaning}\\
\midrule
$d$ & shared latent / embedding dimension\\
$\Phi$ & frozen scenario encoder, $\Phi:\text{scenario}\to\R^d$\\
$D$ & semantic divergence, Eq.~\eqref{eq:divergence}\\
$\tau_{\uparrow},\tau_{\downarrow}$ & upper/lower mitosis thresholds (hysteresis)\\
$k^\star$ & selected number of children\\
$\mathbf{M}\in\R^{S\times d}$ & hormonal bus (shared latent field), $S$ slots\\
$a_j$ & cosine-attention weight over slot $j$, Eq.~\eqref{eq:attn}\\
$E_i$ & energy state of cell $i$, Eq.~\eqref{eq:energy}\\
$\kappa,\eta,\lambda_{\text{basal}}$ & uncertainty, regeneration, basal-decay coefficients\\
$H_{\text{norm}}$ & normalized Shannon uncertainty of the output\\
$\mathbf{P},\mathbf{C},\gamma$ & prototypes, coupling matrix, coupling strength\\
$\mathbf{S}^\star$ & analytic ground truth, Eq.~\eqref{eq:groundtruth}\\
$N$ & declared number of births (CSR denominator)\\
$1-\alpha$ & Wilson confidence level for CSR\\
$R$ & replicate / repetition count\\
\bottomrule
\end{tabular}
\end{table}

\section*{Manuscript Status}
\noindent This is a Stage-1 (design and protocol) registered report. It contains
no empirical results by design. A Stage-2 report will present the executed
protocol and its outcomes, including any refutation of the stated hypotheses.

\end{document}
